\section{Related and Future Work}
\label{sec:related-work}

Using the Coq proof assistant as a uniform platform for implementation
of logic-based program verification tools is a well-established
approach, which by now has been successfully employed in a number of
projects on certified compilers~\cite{Leroy:POPL06,Appel-al:BOOK14}
and verified low-level
code~\cite{Shao:CACM10,Chlipala:PLDI11,Jensen-al:POPL13}, although,
with no specific focus on abstractions for fine-grained concurrency,
such as protocols and auxiliary state.
%
% Our FCSL development inherits a lot of ideas from works on
% verification of sequential imperative programs,
% e.g.,~\cite{Nanevski-al:ICFP08,Nanevski-al:POPL10}.

{
\setlength{\belowcaptionskip}{-10pt}  
\begin{table}
{
%\footnotesize
\sffamily\small % tabular data either 10pt times, or 9pt helvetica
\centering
\begin{tabular}{@{}l@{\ \ \ }c@{\ \ \ }c@{\ \ }c@{\ }c@{\!\!\!}c@{\!}c@{\!\!\!\!}c}
  \textbf{Program} &  
  \rot{\cccode{Priv}} & \rot{\cccode{CLock}} & \rot{\cccode{TLock}} &
  \rot{\cccode{ReadPair}} & \rot{\cccode{Treiber}} & \rot{\cccode{SpanTree}}
  & \rot{\cccode{FlatCombine}}   
  \\ \hline 
  CAS-lock & \yep & \yep &&&&&
  \\
  Ticketed lock & \yep && \yep &&&
  \\
  CG increment & \yep & \yepl & \yepl & & & &
  \\
  CG allocator & \yep & \yepl & \yepl &&&&
  \\
  Pair snapshot & & & & \yep &&&
  \\
  Treiber stack & \yep & \yepa & \yepa & & \yep &&
  \\
  Spanning tree & \yep & & & & & \yep &
  \\
  Flat combiner & \yep & \yepa & \yepa & & & & \yep
  \\ 
  Seq. stack & \yep & \yepa & \yepa & & \yep & &
  \\
  FC-stack & \yep & \yepa & \yepa & & & & \yep
  \\
  Prod/Cons & \yep & \yepa & \yepa & & \yep & &
  \\[2pt] \hline
\end{tabular}
%\vspace{-5pt}
\caption{
  % 
  Primitive concurroids (in column headings) employed by different programs. Two lock
  concurroids, for CAS-based and ticketed locks, are interchangeable, as they implement the same abstract interface
  (indicated by \yepl).  
%
}
\label{tab:concur}
}
\end{table}
}
%

\paragraph{Related program logics}
\label{sec:relat-logic-fram}

The FCSL logic has been designed as a generalization of the classical
Concurrent Separation Logic by O'Hearn~\cite{OHearn:TCS07}, combining
the ideas of local concurrent protocols with arbitrary
interference~\cite{Jones:TOPLAS83,Feng:POPL09} and compositional (\ie,
subjective) auxiliary state~\cite{LeyWild-Nanevski:POPL13} with the
possibility to compose protocols.
%
Other modern concurrency logics, close to FCSL in their expressive
power, are iCAP~\cite{Svendsen-Birkedal:ESOP14},
Iris~\cite{Jung-al:POPL15}, CoLoSL~\cite{Raad-al:colosl}, and
CaReSL~\cite{Turon-al:ICFP13}.
 
Both iCAP and Iris leverage the idea, originated by Jacobs and
Piessens~\cite{Jacobs-Piessens:POPL11}, of parametrizing specs for
fine-grained concurrent data types by client-provided auxiliary code,
which can be seen as a ``callback'', and nowadays is usually referred
to as a \emph{view shift}. View shifts also serve purposes similar to
FCSL's concurroid transitions. A form of composition of concurrent
resources can be encoded by combining view shifts with fractional
permissions~\cite{Bornat-al:POPL05} (in iCAP) or invariant masks (in
Iris).
%
Similarly to FCSL, Iris makes use of PCMs as a generic mechanism to
describe state, which can be split between parallel threads.
%
However, neither iCAP, nor Iris have explicit subjective dichotomy of
the auxiliary state, which makes encoding of thread-specific
contributions (\eg, marked nodes from the graph example) in them less
direct comparing to FCSL.

CoLoSL defines a different notion of thread-local views to a shared
resource, and uses \emph{overlapping
  conjunction}~\cite{Hobor-Villard:POPL13} to reconcile the
permissions and capabilities, residing in the shared state between
parallel threads.
%
Overlapping conjunction affords a description of the shared structure
mirroring the recursive calls in the structure's methods. In FCSL,
such machinery isn't required, as \emph{self} and \emph{other} suffice
to represent the thread-specific views, and \emph{joint} state doesn't
need to be divided.
%
In our opinion, this leads to simpler specs and proofs.
%
For example, the proof that \code{span} constructs a \emph{tree} in
CoLoSL involves abstractions such as shared capabilities for marking
nodes and extension of the graph with virtual edges, none of which is
required in FCSL.
%
Moreover, CoLoSL doesn't prove that the tree is spanning, which we
achieve in FCSL via hiding.

% \begin{comment}
% CoLoSL defines an alternative notion of a thread-specific view to a
% shared resource, defined \wrt the real heap (in contrast to FCSL's
% subjective views, which are parametrized over a generic PCM structure).  CoLoSL
% makes use of the \emph{overlapping
%   conjunction}~\cite{Hobor-Villard:POPL13} to reconcile ``local''
% views of several parallel threads---an idea, generalized by FCSL's
% notion of subjectivity.
% %
% CoLoSL was used to verify the concurrent spanning tree algorithm,
% proving that its result is \emph{a tree}. It was done via logical
% action capabilities and instrumenting the graph with virtual edges.
% %
% However, the CoLoSL paper doesn't prove the resulting tree to be a
% spanning one, which we achieve in FCSL via hiding.
% %
% % \ab{Should the above be: However, the CoLoSL paper doesn't establish
% % that the resulting tree is a \emph{spanning} tree, which we achieve
% % in FCSL using hiding.}
% \end{comment} 

CaReSL combines the Hoare-style reasoning and proofs about
\emph{contextual refinement}.  Similarly to FCSL, CaReSL employs
resource protocols to specify thread interference, although, targeting
the ``life stories'' of particular memory locations instead of
describing a whole concurrent data structure by means of an STS.
While FCSL is not equipped with abstractions for contextual
refinement, in our experience it was never required to prove the
desired Hoare-style specs for fine-grained data structures.

Neither of the four mentioned logics features hiding as a lanugage
constructor for controlling interference.
%
Reasoning in iCAP, Iris, CoLoSL and CaReSL follows the tradition of
Hoare-style logics, so the specs never mention explicitly the heap and
state components. In contrast, FCSL assertions use explicit variables
to bind heap and auxiliary state, as well as their components. In our
experience, working directly with the state model is pleasant, and has
to be done in Coq anyway, since Coq lacks support for contexts of
bunched implications, as argued by
Nanevski~\etal~\cite{Nanevski-al:POPL10}.
%
None of iCAP, CoLoSL or CaReSL features a mechanized metatheory, and
neither of these logics has been implemented as a mechanized
verification tool. The soundness of Iris' primitive proof rules has
been mechanized in Coq, however, the logic itself has only been
employed for paper-and-pencil verification of example programs.

\paragraph{Related tools for concurrency verification}
%
SAGL and RGSep, the first logics for modular reasoning about
fine-grained
concurrency~\cite{Feng-al:ESOP07,Vafeiadis-Parkinson:CONCUR07},
inspired creation of semi- and fully-automated verification tools:
SmallfootRG~\cite{Calcagno-al:SAS07} and
Cave~\cite{Vafeiadis:VMCAI10}.
%
These tools target basic safety properties of first-order code,  such
as data integrity and absence of memory leaks.

Chalice~\cite{Leino-Muller:ESOP09} is an experimental first-order
concurrent language, supplied with a tool that generates
verification conditions (VCs) for client-annotated Chalice
programs. Such VCs are suitable for automatic discharge by SMT
solvers.
%
% Chalice has been recently employed as a backend in a number of
% verification frameworks for concurrency, such as
% VerCors~\cite{Blom-Huisman:FM14}.
%
For local reasoning, Chalice employs fractional
permissions~\cite{Bornat-al:POPL05}, implicit dynamic frames, and
auxiliary state~\cite{Leino-al:FOSAD09}, which, unlike the one of
FCSL, is not a subject of PCM laws, and thus is not compositional, as
its shape should match the forking pattern of the client program being
verified.
%
Chalice also supports a form of \emph{symmetric} Rely/Guarantee
reasoning, \ie, it does not allow the threads to take different roles
in a protocol (which is expressible in FCSL via \emph{self}-enabled
transitions, such as \code{nullify_trans} from
Section~\ref{sec:graph-conc-coher}).
%
Chalice's specification fragment is a first-order logic, whereas FCSL
admits higher-order functions and predicates in specs, therefore,
enabling program composition and proof reuse, as shown in
Figure~\ref{fig:deps}.

VCC~\cite{Cohen-al:TPHOL09} is a tool for verifying low-level
concurrent C code.  VCC relies on an object-based model to describe
protocols and decomposition of state, and its two-state object
invariants are similar to our resource protocols. However precise
comparison is difficult, as FCSL is not bound to an object-based model
and allows quantification over arbitrary state. For compositional
reasoning, VCC employs a number of abstractions, such as
\emph{claims}, \emph{ghost} and \emph{volatile} fields, \emph{ghost},
\emph{owned}, \emph{wrapped}, \emph{nested} and \emph{closed}
objects. Because FCSL relies on standard type theory and PCMs,
specialized abstractions and annotations are not required. VCC allows
specifications only in a first-order logic to support automation for
an SMT-based back-end. To the best of our knowledge, soundness of full
VCC has not been proved formally~\cite{Cohen-al:SOFSEM13}.

VeriFast~\cite{Jacobs-al:NFM11} is a tool for deductive verification
of sequential and concurrent C and Java programs, based on separation
logic~\cite{Reynolds:LICS02}.
% 
%
To specify and verify fine-grained concurrent algorithms, VeriFast
employs fractional permissions~\cite{Bornat-al:POPL05} and a form of
first-class auxiliary code~\cite{Jacobs-Piessens:POPL11}, enabling
manipulation of (non-compositional) auxiliary state. VeriFast has been
recently extended with Rely/Guarantee
reasoning~\cite{Jacobs-al:sharedboxes}, although, without a
possibility to compose resources. To the best of our knowledge,
soundness of VeriFast core has been formally proved only
partially~\cite{Vogels:PhD}.

% Other shallow-embedding-based approaches~\cite{Brady-Hammond:FI10}

% Several projects has been conducted recently by means of \emph{deep}
% embedding (\ie, by implementing a standalone language) of the XCAP
% framework into
% Coq~\cite{Chlipala:PLDI11,Feng-al:PLDI08,Ni-Shao:POPL06}, targeting
% verification of different aspects of low-level code, without
% introducing specific abstractions for fine-grained concurrency.

% \citeyearpar{Gordon:PhD}

Rely-Guarantee references (\textsc{RGRef}s) by Gordon~\cite{Gordon:PhD} are
a mechanism to prove transition invariants of concurrent data
structures.
%
%
While the language of \textsc{RGRef}s is implemented as an axiomatized
Coq DSL, the system's soundness is proved by hand using the Views
framework~\cite{DinsdaleYoung-al:POPL13}.
%
Since \textsc{RGRef}s focuses on correctness of data structures \wrt
specific protocols and doesn't provide auxiliary state, it's unclear
how to employ it for client-side reasoning.

\paragraph{Future work}
\label{sec:future-work}
%
In the future, we plan to augment FCSL with the program extraction
mechanism~\cite{Letouzey:CIE08} and implement proof automation for
stability-related facts via lemma
overloading~\cite{Gonthier-al:ICFP11}.
%
Due to the limitations of Coq's model \wrt impredicativity, at this
moment, FCSL doesn't support higher-order heaps (\ie, the possibility
to reason about arbitrary storable effectful procedures). While programs
requiring this feature are rare, we hope that it will be possible to
encode and verify the characteristic ones, once the development is
migrated to Coq 8.5, featuring universe
polymorphism~\cite{Sozeau-Tabareau:ITP14}.
