\documentclass[runningheads]{llncs}

%% Save the class definition of \subparagraph
\let\llncssubparagraph\subparagraph
%% Provide a definition to \subparagraph to keep titlesec happy
\let\subparagraph\paragraph
%% Load titlesec
\usepackage[compact]{titlesec}
%% Revert \subparagraph to the llncs definition
\let\subparagraph\llncssubparagraph

\usepackage{amssymb,amsmath}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{txfonts} %% for various points-to symbols. use pxfonts for 5-tip stars
\usepackage{verbatim} %% for comment environment
\usepackage{xspace} %% for comment environment
\usepackage{cancel} % for crossing Owns 
\usepackage{mathpartir}
\usepackage{soul}
\usepackage{xcolor}
\usepackage{multirow}
\usepackage{stmaryrd} 
% % sorting citation
\usepackage{url}
\usepackage{float} 
\usepackage{framed}
\usepackage{hyperref}
% Fonts
\usepackage{bbm}
\usepackage{alltt}
\usepackage{verbdef}
\usepackage{titlesec}
\usepackage{parskip}
\usepackage{wrapfig}
\usepackage{cite}

\setlength{\intextsep}{5pt}%

\setlength{\parindent}{0.15in}
\setlength{\topsep}{0cm}
\setlength{\parskip}{0pt}
\titlespacing*{\section}{0pt}{*1}{*1}
\titlespacing*{\subsection}{0pt}{*0.7}{*0.5}
\titlespacing*{\paragraph}{0pt}{*0.5}{*0.5}
\renewcommand{\figurename}{Fig.}

\hyphenation{time-stamp time-stamps}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Floats

\renewcommand{\textfraction}{0.1}
\renewcommand{\topfraction}{0.95}
\renewcommand{\dbltopfraction}{0.95}
\renewcommand{\floatpagefraction}{0.9}
\renewcommand{\dblfloatpagefraction}{0.9}

\setlength{\floatsep}{16pt plus 4pt minus 4pt}
\setlength{\textfloatsep}{16pt plus 4pt minus 4pt}

\floatstyle{ruled} 
\restylefloat{figure}  

% reducing the sf math fonts somewhat
\usepackage[T1]{fontenc}

\DeclareFontFamily{OML}{zavm}{\skewchar\font=127 }
\DeclareFontShape{OML}{zavm}{m}{n}{<-> s*[.80] zavmr7t}{}
\DeclareFontShape{OML}{zavm}{b}{n}{<-> s*[.80] zavmb7t}{}
\DeclareFontShape{OML}{zavm}{m}{it}{<-> s*[.80] zavmri7m}{}
\DeclareFontShape{OML}{zavm}{b}{it}{<-> s*[.80] zavmbi7m}{}
\DeclareFontShape{OML}{zavm}{m}{sl}{<->ssub * zavm/m/it}{}
\DeclareFontShape{OML}{zavm}{bx}{it}{<->ssub * zavm/b/it}{}
\DeclareFontShape{OML}{zavm}{b}{sl}{<->ssub * zavm/b/it}{}
\DeclareFontShape{OML}{zavm}{bx}{sl}{<->ssub * zavm/b/sl}{}
\DeclareMathAlphabet{\mathsf}{OML}{zavm}{m}{n} % not `n'

% Figures should be boxed
%    *** Uncomment the next two lines to box the floats *** 
% \floatstyle{boxed}
% \restylefloat{figure}

% Keep footnotes on one page
\interfootnotelinepenalty=10000 

\input{defs}
\newcounter{tags}

\newcommand{\mytitle}{Specifying and Verifying Concurrent
  Algorithms \\
with Histories and Subjectivity}
%\newcommand{\mytitle}{
%  Specifying and Verifying Fine-Grained Concurrent Algorithms \\{} 
%  using Histories and Subjectivity}
% \newcommand{\mytitle}{
%   Specifying Interference in Hoare-Style Logic for Concurrency}

\clubpenalty=10000 
\widowpenalty = 10000 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Floats

\begin{document} 

%\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

% \subtitle{Extended version}
%\subtitle{\large{\color{red}\today}}

%

\author{Ilya Sergey \and 
  Aleksandar Nanevski \and
  Anindya Banerjee}

\institute{IMDEA Software Institute, Spain\\ 
\email{\{ilya.sergey, aleks.nanevski, anindya.banerjee\}@imdea.org}}

\title{\mytitle}

\maketitle 

\begin{abstract}

  We present a lightweight approach to Hoare-style specifications for
  fine-grained concurrency, based on a notion of \emph{time-stamped
    histories} that abstractly capture atomic changes in the program
  state.
% 
Our key observation is that histories form a \emph{partial commutative
  monoid}, a structure fundamental for representation of concurrent
resources.
%
This insight provides us with a unifying mechanism that allows us to
treat histories just like heaps in separation logic. For example, both
are subject to the same assertion logic and inference rules (\eg, the
frame rule). Moreover, the notion of ownership transfer, which usually
applies to heaps, has an equivalent in histories. It can be used to
formally represent helping---an important design pattern for
concurrent algorithms whereby one thread can execute code on behalf of
another.
%
Specifications in terms of histories naturally abstract away the
internal interference, so that sophisticated fine-grained algorithms
can be given the same specifications as their simplified
coarse-grained counterparts, making them equally convenient for
client-side reasoning.
% 
We illustrate our approach on a number of examples and validate all of
them in~Coq.
  
\end{abstract}

% \category{D.1.1}{Programming Techniques}{Applicative (Functional) Programming}
% %
% \category{F.3.2}{Logics and Meanings of Programs}{Semantics of
%  Programming Languages --- Program analysis, Operational semantics}

% do nothing

% \terms
% Languages, Theory, Verification

\input{intro5}
\input{overview2}
\input{background2}
\input{examples2}
\input{flatcombine2}
\input{related}
\input{conclusion}
 
%\setlength{\bibsep}{2.2pt}

\bibliographystyle{abbrv}
\bibliography{bibmacros,references,proceedings-short}

% \newpage
% \appendix
% \input{appendix}

\end{document}

