\section{Background on FCSL}
\label{sec:background}

A Hoare specification in FCSL has the form
$\spec{P}\ e\ \spec{Q} @ \rcon$. $P$ and $Q$ are the precondition and
postcondition for $e$, and $\rcon$ defines the \emph{resource} on
which $e$ operates. We have elided $\rcon$ from the specs in
Section~\ref{sec:overview}, but explain it now. A resource is a state
transition system describing the state (real and auxiliary) and atomic
operations that the threads that want to simultaneously operate on
that state have to respect. For example, the state space
(aka.~resource invariants) of the resource $\cal E$ for the exchanger
is described by the predicates~\eqref{tag:exchanging} and
\ref{exP}--\ref{ex:gapless}. On the other hand, the transitions of
$\cal E$ can be read off from our discussion of the exchanger proof
outline, where at each atomic operation of \code{CAS}, we explicitly
described how the operation modifies the auxiliary state. These
modifications are the transitions that one has to declare in $\rcon$,
and then prove, when establishing $\spec{P}\ e\ \spec{Q} @ \rcon$,
that $e$ only makes declared transitions.

But a resource also has an important secondary role. Its definition
provides the variables that $P$ and $Q$ may scope over. For example,
in the case of exchanger, we used the variables $\heaps, \perms,
\hists$, $\heapo, \permo, \histo$, and $\heapj, \pending$.  Of course,
different resources will provide different state spaces, transitions
and variables. For example, a commonly used resource is $\cal P$ for
\emph{private state}. $\cal P$ allows only variables $\heaps$ and
$\heapo$ of type heap, denoting the private heap of a thread, and the
private heap of other threads. The transitions of $\cal P$ allow for
reading, writing, allocating and deallocating pointers from $\heaps$.

The mechanism by which a resource defines the allowed variables is as
follows. Underneath, a resource comes with only three variables:
$a_\lcl$, $a_\env$ and $a_\joint$ standing for abstract self state,
other state, and shared (joint) state, but the user can pick their
types depending on the application. In the case of $\cal E$, $a_\lcl$
and $a_\env$ are triples containing a heap, an offer-set and a
history. The variables we used in Section~\ref{sec:overview} are then
merely projections out of such triples:
$a_\lcl\,{=}\,(\heaps, \perms, \hists)$, and
$a_\env\,{=}\,(\heapo, \permo, \histo)$. Similarly,
$a_\joint\,{=}\,(\heapj, \pending)$.

It is essential that $a_\lcl$ and $a_\env$ have a common type, which
moreover, exhibits the algebraic structure of a \emph{partial
  commutative monoid} (PCM). A PCM requires a partial binary operation
$\bullet$ which is commutative and associative, and has a unit. In the
case of $\cal E$, each of the three components of $a_\lcl$ and
$a_\env$---heaps, offer-sets and histories---form a PCM, where
$\bullet$ is disjoint union $\hunion$, and $\emptyset$ is the
unit. Hence, the product of the three is a PCM as well, with $\bullet$
and unit defined point-wise. PCMs are important, as they give a way,
generic in $\rcon$, to define the inference rule for parallel composition.
%
\[
\tag{\normalsize \arabic{tags}}\refstepcounter{tags}\label{eq:parrule}
{\small{
\begin{array}{c}
\specK{\{P_1\}}\ e_1\ \specK{\{Q_1\}} @ \rcon \quad \specK{\{P_2\}}\ e_2\ \specK{\{Q_2\}} @ \rcon\\[2pt]
\hline\\[-7pt]
\specK{\{P_1 \circledast P_2\}}\ e_1 \parallel e_2\ \specK{\{[\res.1/\res]Q_1 \circledast [\res.2/\res]Q_2\}} @ \rcon
\end{array}
}}
\]
%
Here, $\circledast$ is defined as follows.
\[
\tag{\normalsize \arabic{tags}}\refstepcounter{tags}\label{eq:ssep}
\begin{array}{c}
(P_1 \circledast P_2)(a_\lcl, a_\joint, a_\env) \iff \exists x_1~x_2\ldot a_\lcl = x_1 \bullet x_2, \hbox{}\\
 P_1 (x_1, a_\joint, x_2 \bullet a_\env), P_2 (x_2, a_\joint, x_1 \bullet a_\env)
\end{array}
\]
%
%
The inference rule, and the definition of $\circledast$, formalize the
intuition that when a parent thread forks $e_1$ and $e_2$, then $e_1$
is part of the environment for $e_2$ and vice-versa. This is so
because the \emph{self} component $a_\lcl$ of the parent thread is
split into $x_1$ and $x_2$; $x_1$ and $x_2$ become the \emph{self}
parts of $e_1$, and $e_2$ respectively, but $x_2$ is also added to the
\emph{other} component $a_\env$ of $e_1$, and dually, $x_1$ is added
to the \emph{other} component of $e_2$.
%
Also note that parallel composition returns a pair of the outputs
produced by $e_1$ and $e_2$. Thus, the variable $\res$ in $Q_1$ and
$Q_2$ has to be appropriately renamed by the projections $\res.1$ and
$\res.2$ in the postcondition of the parallel composition.

The rule of frame of FCSL is a special case of parallel composition,
when $e_2$ is the idle thread.
%
\[
\tag{\normalsize \arabic{tags}}\refstepcounter{tags}\label{eq:frame}
{\small{
\begin{array}{c}
\specK{\{P_1\}}\ e\ \specK{\{P_2\}} @ \rcon\\[2pt]
\hline\\[-7pt]
\specK{\{P_1 \circledast Q\}}\ e\ \specK{\{P_1 \circledast Q\}} @ \rcon
\end{array}\qquad 
\begin{array}{c}
\mbox{$Q$ stable under}\\
\mbox{$\rcon$'s transitions}
\end{array}
}}
\]
A notable difference from the frame rules of other separation logics
is that FCSL's definition of $\circledast$ forces that the value being
framed onto \emph{self} component is \emph{subtracted} from the
\emph{other} component, whereas in other separation logic, the frame
value materializes out of nowhere. To illustrate, we can frame
$\gists$ onto the history $\hists$ in the the
spec~(\ref{tag:exchangespec}), by taking
$\rcon\,{\eqdef}\,a_\lcl\,{=}\,(\heaps, \hists,
\perms)\,{=}\,(\emptyset, \gists, \emptyset)$.
We obtain, after some simplification:
%
\[
{\small{
\begin{array}{c}
\specK{\{\heaps = \emptyset, \perms = \emptyset, \hists = \gists, \gist \subseteq \gists \hunion \histo \hunion \mygather{\pending}\}}\\[2pt]
\mathtt{exchange}\ v \\[2pt]
\spec{\!\!
  \begin{array}{c}
    \heaps = \emptyset, \perms = \emptyset, \gist \subseteq \gists \hunion \histo \hunion \mygather{\pending}, \hbox{}\\[1pt]
    \mathsf{if}\ \res\ \mathsf{is}\ \mathsf{Some}\ w\ \mathsf{then}\\[1pt]
    \exists t\ldot \hists = t \mapsto (v, w) \hunion \gists, 
    \mathsf{last} (\gist) < t, \twin{t}~\mathsf{else}\ \hists = \gists    
  \end{array}
\!\!}@\cal E
\end{array}
}}
\]
But notice how the spec now says that $\gist \subseteq \gists \hunion
\histo \hunion \mygather{\pending}$, whereas
in~(\ref{tag:exchangespec}) it said $\gist \subseteq \histo \hunion
\mygather{\pending}$. The addition of $\gists$ compensates for
$\gists$ having been subtracted out of $\histo$, to be moved to
$\hists$.

Finally, we will use one more constructor of FCSL, and its associated
inference rule: \emph{hiding}. The program $\mathsf{hide}\ e$
operationally just executes $e$, but logically allows installing a
resource within the scope of $e$. For example, a program can start
only with private heap variables $\heaps$ and $\heapo$ (\ie, using
only the resource $\cal P$). However, it can then take a chunk of heap
out of $\heaps$ and ``install'', for instance, an exchanger resource
$\cal E$ in it, thereby giving the threads in $e$ the ability to
exchange values. The spec of $e$ is then expressed using the variables
that $\cal E$ adds to $\cal P$: $\perms, \hists$, $\permo, \histo$,
and $\heapj, \pending$. Upon termination, the extra variables are
removed from the scope. We elide here the general form of the hiding
rule (it is given in~\cite{Nanevski-al:ESOP14}), and just provide the
special case involving $\cal E$ and $\cal P$.
\[
\tag{\normalsize \arabic{tags}}\refstepcounter{tags}\label{eq:ehide}
{\small{
\begin{array}{c}
\specK{\{P\}}\ e\ \specK{\{Q\}} @ \cal E\\[2pt]
\hline\\[-7pt]
\specK{\{\heaps = \Phi_1(\heapj), \Phi_1(P)\}}\ \mathsf{hide}_{\Phi_1}~e\ \specK{\{\exists \Phi_2\ldot \heaps = \Phi_2(\heapj), \Phi_2(Q)\}} @ \cal P
\end{array}
}}
\]
When read bottom up, the rule says that we can install the resource
$\cal E$ in the scope of a thread that works with $\cal P$, but then
we need substitutions $\Phi_1$ and $\Phi_2$, to map variables of $\cal
E$ ($\heaps, \perms, \hists$, \etc) to values expressed with variables
from $\cal P$ ($\heaps$, and $\heapo$). Here $\Phi_1$ is an initial
such substitution (user provided), and the rule guarantees the
existence of an ending substitution $\Phi_2$. The substitutions have
to satisfy a number of side conditions, which we elide here for
brevity. The most important one is that $\Phi_1$, $\Phi_2$, fix the
\emph{other} variable $a_\env = (\heapo, \permo, \histo)$ to be the
PCM unit (\ie,~a triple of empty sets). Fixing $a_\env$ to unit
formalizes the intuition that within the scope of $\mathsf{hide}$,
the outside threads cannot see $\cal E$, and thus cannot interfere
with the exchanges performed by $e$. They start with empty heap,
offer-set and history, and terminate with empty ones too.

At the beginning of $\mathsf{hide}~e$, the private heap equals the
value that $\Phi_1$ gives to $\heapj$ ($\heaps = \Phi_1(\heapj)$). In
other words, the $\mathsf{hide}$ rule takes the private heap of a
thread, and makes it shared, \ie, gives it to the $\heapj$ component
of $\cal E$. Upon finishing, $\mathsf{hide}~e$ makes $\heapj$ private
again.
%
%($\heaps = \Phi_2(\heapj)$).

%\an{Should I say something about compositionality? Why is FCSL
%  compositional? Maybe say, soundness of FCSL has been established by
%  shallow embeding in Coq. Thus, the logic immediately inherits the
%  substitution principle, thereby allowing that clients can reason
%  only out of the Hoare spec of an object.}

%\subsection{FCSL basics}
%\label{sec:fcsl-basics}
%
%\todo{A short overview of FCSL: mostly, concerning subjectivity and hiding}
%
%\subsection{Histories as auxiliary state}
%\label{sec:hist-state} 

