\section{Related work and discussion}
\label{sc:related}

% Early times

\paragraph{Program logics for linearizability.}

The proof method for establishing linearizability of concurrent
objects based on the notion of linearization points has been
presented in the original paper by Herlihy and
Wing~\cite{HerlihyW+TOPLAS90}. The first Hoare-style logic, employing
this method for compositional proofs of linearizability was introduced
in Vafeiadis' PhD thesis~\cite{Vafeiadis+PhD07}. However, that logic,
while being inspired by the combination of Rely-Guarantee reasoning
and Concurrent Separation logic~\cite{VafeiadisP+CONCUR07} with
syntactic treatment of linearization
points~\cite{VafeiadisHHS+PPoPP06}, did not have a soundness proof
with respect to any program semantics. Furthermore, the
work~\cite{Vafeiadis+PhD07} did not connect reasoning about
linearizability to the verification of client programs that make use
of linearizable objects in a concurrent environment (\cf
Section~\ref{sc:clients}).

% Modern logics for linearizability

Both these shortcomings were addressed in more recent works on program
logics for linearizability~\cite{LiangF+PLDI13,Khyzha-al:FM16}, or,
equivalently, \emph{observational
  refinement}~\cite{FilipovicOHRW+TCS10,TuronDB+ICFP13}. These works
provided semantically sound methodologies for verifying refinement of
concurrent objects, by encoding atomic commands as resources
(sometimes encoded via a more general notion of
\emph{tokens}~\cite{Khyzha-al:FM16}) directly into a Hoare
logic. Moreover, the logics~\cite{LiangF+PLDI13, TuronDB+ICFP13}
allowed one to give the objects standard Hoare-style specifications.
%
However, in the works~\cite{LiangF+PLDI13,TuronDB+ICFP13}, these two
properties (\ie,~linearizability of a data structure and validity of
its Hoare-style spec) are established separately, thus doubling the
proving effort.
%
That is, in those logics, provided a proof of linearizability for a
concurrent data structure, manifested by a spec that suitably handles
a \emph{command-as-resource}, one should then devise a declarative
specification that exhibits temporal and spatial aspects of executions
(akin to our history-based specs from Figure~\ref{fig:specs}),
required for verifying the client code.

Importantly, in those logics, determining the linearization order of a
procedure is tied with that procedure ``running'' the
command-as-resource within its execution span. This makes it difficult
to verify programs where the procedure terminates before the order is
decided on, such as \jywrite~operation in Jayanti's snapshot. The
problem may be overcome by extending the scope of \emph{prophecy
  variables}~\cite{AbadiL+lics88} or \emph{speculations} beyond the
body of the specified procedure. However, to the best of our
knowledge, this has not been done yet.

% \an{Jung told
%   German at ICFP that \cite{TuronTABD+POPL13} has an example of a
%   conditional CAS, which is non-regional. And indeed, they explicitly
%   say on page 2, first column, that they support reasoning about
%   temporally-non-local linearization points, which is their word for
%   non-regional. They say they do it via speculation!!! I was up to now
%   under the impression that speculations can handle future-dependent
%   linearization points, but only by the time the procedure
%   terminates. So, I'm puzzled; I guess different people mean different
%   things under ``speculation''. That said, this paper is the model
%   behind CaReSL, not the CaReSL logic itself. Nobody's using CaReSL
%   now, which says that something is off in it regarding the
%   interaction between logical refinement (\ie, linearization) and
%   Hoare stuff. This may explain why IRIS people gave talks at Dagstuhl
%   saying that speculations are a problem for them.  Also, did I
%   mischaracterize the prophecy variables business here? In the
%   previous version, these two logics were not tied to
%   prophecies. Should they have been?}  \an{At any rate, we may
%   consider weakening or removing the statement that these people can't
%   do non-regional. Instead, we may just say that they prove
%   refinement, but double the proving effort when it comes to combining
%   that with Hoare logic specs, whereas we just do it all at once,
%   which has its advantages, as we already discussed a bit in Section
%   3.}

%
% No-linearizability

\paragraph{Hoare-style specifications as an alternative to
  linearizability.} 

A series of recent Hoare logics focus on specifying concurrent
behavior \emph{without} resorting to
linearizability~\cite{SergeyNB+ESOP15, SergeyNBD+OOPSLA16, SvendsenB+ESOP14,
  PintoDYG+ECOOP14, JungSSSTBD+POPL15}.
%
This paper continues the same line of thinking, building
on~\cite{SergeyNB+ESOP15}, which explored patterns of assigning
Hoare-style specifications with self/other auxiliary histories to
concurrent objects, including \emph{higher-order} ones (e.g., {flat
  combiner}~\cite{HendlerHIST+SPAA10}), and \emph{non-linearizable}
ones~\cite{SergeyNBD+OOPSLA16} in FCSL~\cite{NanevskiLSD+ESOP14}, but
has not considered non-local, future-dependent linearization points,
as required by Jayanti's algorithm.
%

% Following the style of specifications, advocated by 
Alternative logics, such as
Iris~\cite{JungSSSTBD+POPL15,JungKBD+ICFP16} and
iCAP~\cite{SvendsenB+ESOP14}, employ the idea of ``ghost
callbacks''~\cite{Jacobs-Piessens+POPL11}, to identify precisely the
point in code when the callback should be invoked.  Such a program
point essentially corresponds to a local linearization
point. Similarly to the logical linearizability proofs, in the
presence of future-dependent LPs, this method would require
speculating about possible future execution of the callback, just as
commented above, but that requires changes to these logics'
metatheory, in order to support speculations, that have not been
carried out yet.

The specification style of TaDA logic~\cite{PintoDYG+ECOOP14} is
closer to ours in the sense that it employs \emph{atomic tracking
  resources}, that are reminiscent of our history entries. However,
the metatheory of TaDA does not support ownership transfer, which is
crucial for verifying algorithms with non-local linearization
points. As demonstrated by this paper and also previous
works~\cite{SergeyNB+ESOP15,SergeyNBD+OOPSLA16}, history entries can
be subject to ownership transfer, just like any other
resources.


The key novelty of the current work with respect to previous results
on Hoare logics with histories~\cite{FuLFSZ+CONCUR10, LiangF+PLDI13,
  GotsmanRY+ESOP13, BellAW+SAS10, SergeyNB+ESOP15, HemedRV+DISC15} is
the idea of representing logical histories as auxiliary state, thus
enabling constructive reasoning, by \emph{relinking}, about
dynamically changing linearization points.
%
Since relinking is just a manipulation of otherwise standard auxiliary
state, we were able to use FCSL \emph{off the shelf}, with no
extensions to its metatheory. Furthermore, we expect to be able to use
FCSL's higher-order features to reason about higher-order (\ie,
parametrized by another data structure) snapshot-based
constructions~\cite{PetrankT+DISC13}.
%
Related to our result, O'Hearn \etal have shown how to employ
history-based reasoning and Hoare-style logic to
\emph{non-constructively} prove the existence of linearization points
for concurrent objects out of the data structure
invariants~\cite{OHearnRVYY+PODC10}; this result is known as \emph{the
  Hindsight Lemma}. The reasoning principle presented in this paper
generalizes that idea, since the Hindsight Lemma is only applicable to
``pure'' concurrent methods (\eg, a concurrent set's
\texttt{contains}~\cite{HellerHLMSS+OPODIS05}) that do not influence
the position of other threads' linearization points. In contrast, our
history relinking handles such cases, as showcased by Jayanti's
construction, where the linearization point of \texttt{write} depends
on the (future) outcome of \texttt{scan}.

%% Aspects Oriented Linearizability and other approeaches to cope with
%% non-local linearization points.

\paragraph{Semantic proofs of linearizability.}
\label{sec:semant-proofs-line}

There has been a long line of research on establishing linearizability
using forward-backwards
simulations~\cite{SchellhornWD+CAV12,ColvinGLM06,ColvinDG05}. These
proofs usually require a complex simulation argument and are not
modular, because they require reasoning about the entire data structure
implementation, with all its methods, as a monolithic
state-transition system.

Recent works~\cite{HenzingerSV+CONCUR13, ChakrabortyHSV+LMCS15,
  DoddsHK+POPL15} describe methods for establishing linearizability of
sophisticated implementations (such as Herlihy-Wing
queue~\cite{HerlihyW+TOPLAS90} or the time-stamped
stack~\cite{DoddsHK+POPL15}) in a modular way, via
\emph{aspect-oriented} proofs.
%
This methodology requires devising, for each class of objects (\eg,
queues or stacks), a set of specification-specific conditions (aka
aspects) characterizing the observed executions, and then showing that
establishing such properties implies its linearizability.
%
This approach circumvents the challenge of reasoning about
future-dependent linearization points, at the expense of (a)
developing suitable aspects for each new data structure class and
proving the corresponding ``aspect theorem'', and (b) verifying the
aspects for a specific implementation. 
%
% \ab{Very long sentence. Can we break it up?}
%
Even though some of the aspects have been mechanized and proved
adequate~\cite{DoddsHK+POPL15}, 
%at the moment, aspect-oriented
%verification of implementations is still a manual effort.
%
currently, we are not aware of such aspects for snapshots.

%  establish the linearizability of Herlihy-Wing
% queue---another algorithm where the linearization order of an event
% may be determined only after the event terminated---but using
% automated verification methods. In contrast, we used a logic for
% modularly establishing full functional specifications, which, while
% not exactly linearizability, in the case of Jayanti's algorithm
% achieves the same goals. \an{Somebody please check that I didn't
%   mischaracterize these papers.}

%... \gad{Complete!!}.  Schellhorn \etal~\cite{SchellhornWD+CAV12},
%... \gad{Complete!!}
%
%\gad{Not sure if this final paragraph belongs here, now}
%% Jayanti's ``Proof''

Our approach is based on program logics and the use of STSs to describe the
state-space of concurrent objects. Modular reasoning is achieved by
means of separately proving properties of specific STS transitions,
and then establishing specifications of programs, composed out of
well-defined atomic commands, following the transitions, and respecting
the STS invariants.

\paragraph{Relation to Jayanti's original proof.}
\label{sec:relat-jayant-orig}

We note that our proof of Jayanti's algorithm seems very different
from Jayanti's original proof. Jayanti relies on so-called
\emph{forwarding principles}, as a key property of the proof. For
example, Jayanti's First Forwarding Principle says (in paraphrase)
that if {\tt scan} misses the value of a concurrent write through
lines~\lineScanReadsX--\lineScanReadsY\ of
Figure~\ref{fig:jayanti-snapshot}, but the write terminates before the
scanner goes through line~\lineScanUnsetsS\ (the scan's linearization
point), then the scanner will catch the value in the forwarding
pointers through lines~\lineScanReadsFX--\lineScanReadsFY.
%
Instead of forwarding principles, we rely on colors to algorithmically
construct the status of each write event as it progresses through
time, and express our assertions using formal logic. For example,
though we did not use the First Forwarding Principle, we nevertheless
can express a similar property, whose proof follows from the auxiliary
state Invariants introduced in Section~\ref{sc:auxiliaries}:
%
\begin{myprop}\label{inv:fwd1}%
If $\sss = \sOff\ \toff$ and $\sx = \sy = \TT$--- i.e., the scanner is
in lines~\lineScanReadsFX--\lineScanReadsFY\ and it has unset $\s$ in
line~\lineScanUnsetsS\ at time $\toff$---then:
%
\[\forall t \in
  \hist\ldot\ t \leq \E(t)< \toff \implies \C(t)= \mathsf{green}\]
\end{myprop}
%
%This proposition is still somewhat different from the forwarding
%principle, because it states that \emph{all} events, not just the
%missed ones, that terminated before the scanner goes through
%line~\lineScanUsetS, are colored green; hence, are ``observed'' by the
%scanner.
%
%Notice that our forwarding principle is a bit stronger than Jayanti's:
%it states that at the moment the scanner reads from the forwarding
%pointers on lines~\lineScanReadsFX--\lineScanReadsFY, all write events
%that finished before \jyscan's line~\lineScanUnsetsS---including those
%originally missed---are green and thus will not be missed by the
%current scan.
%
%\gad{Better names are welcome! Does this exaplanation helps? In fact,
%  the statement reads as if we are verifying that we comply with
%  Jayanti's FFP.}

% \item Kyzha et al. track the ordering of timestamps quite differnetly
%   from us. Where we keep an ``existential'' witness for the total
%   ordering of timestamps, at all stages of evaluation, they do so
%   ``universially''. Thus, they require proving that all possible
%   completions of a partial order into a total order are valid for
%   establishing the relation with the linearization program. We believe
%   this leads to larger proofs and more complicated proofs than
%   necessary.
% \end{enumerate}

\paragraph{Alternative snapshot
  implementations and substitution principle.}

We close this section by discussing our previous
verification of another snapshot algorithm~\cite{SergeyNB+ESOP15}. 

\input{readpair}
%
\noindent
%
We present only \jyscan\ in Figure~\ref{fig:readpair}, as \jywrite\ is
trivial.  In this example, the snapshot structure consists of pointers
$x$ and $y$ storing tuples $(c_x, v_x)$ and $(c_y, v_y)$,
respectively. $c_x$ and $c_y$ are the payload of $x$ and $y$, whereas
$v_x$ and $v_y$ are version numbers, internal to the structure. Writes
to $x$ and $y$ increment the version number, while {\tt scan} reads $x$, $y$
and $x$ again, in succession. Snapshot inconsistency is avoided by
restarting if the two version numbers of $x$ differ. In the notation
of the current paper, the spec we proved is:
\[
\esc{scan} : \tsPos{\{\histS = \hempty\}}\ \tsPos{\{\exists t\ldot \histS' = \hempty \wedge r = \eval\ t\ \histP \wedge \dom{\hist} \subseteq \ideal{\histP}{t}\}}
\]
Indeed the spec is very similar to the one from
Figure~\ref{fig:specs}, but exhibits that the algorithm does not
require dynamic modification to the event ordering. Thus, by defining
$\stableorder$ to be the natural numbers ordering on timestamps in the
global history $\hist$ (so that $\ideal{\stableorderP}{t} =
\ideal{\histP}{t}$), and taking $\scanned{\stableorder}$ to be the set
of all timestamps in $\hist$ (so that $t \in \scanned{\stableorder}$
is trivially true and can be added to the postcondition above), the
above spec directly weakens into that of Figure~\ref{fig:specs}. As
FCSL satisfies a substitution principle whereby client proofs are
developed out of the specs, and not the code of programs, we can
substitute different implementations of snapshot algorithms in
clients, without disturbing the client proof. This is similar in
spirit to the property that programs that linearize to the same
sequential code are interchangeable in clients.
