% config from instructions webpage
\documentclass[a4paper,USenglish]{lipics-v2016}

\usepackage{microtype}%if unwanted, comment out or use option "draft"

\usepackage{authblk}
\usepackage{fancyhdr}
\usepackage{url}
\usepackage{lipsum}
\input{flags}
\input{defs}

\title{Concurrent Data Structures Linked in Time}

\author{Germ\'{a}n Andr\'{e}s Delbianco}
\author{Ilya Sergey}
\author{Aleksandar Nanevski}
\author{Anindya Banerjee}

% \affil[1]{IMDEA Software Institute,  Madrid, Spain\\
%           {\texttt{\{german.delbianco,aleks.nanevski,anindya.banerjee\}@imdea.org}}}

% \affil[2]{University College London, UK\\
%   {\texttt{i.sergey@ucl.ac.uk}}}

% \affil[3]{Universidad Polit\'{e}cnica de Madrid, Spain}

\begin{document}
\maketitle
 
\thispagestyle{empty}


\section*{\LARGE{Note to Reviewers}}
\label{sec:note-reviewers}

This paper has been previously submitted for publication at CONCUR
2016 and ESOP 2017.

\subparagraph{Issues with the CONCUR 2016 submission.}

The submitted manuscript: 

\begin{center}
  \url{https://arxiv.org/pdf/1604.08080v1.pdf}
\end{center}

\noindent
%
The CONCUR~2016 reviewers were concerned about (a) the complexity of
the initial specifications, (b) the incomplete mechanization of the
proof of snapshot specifications, which at that moment relied on
several lemmas, whose proofs were conducted by hand and (c) the lack
of client applications.

All of those issues have been addressed since then. To address (a), we
simplified the specs significantly, so they would exclude irrelevant
implementation-specific constraints on the auxiliary state. For (b),
we have finished the mechanized proofs. Finally, to address (c), we
added a dedicated section describing client applications of the
established specs that are provided in Section~4 of the current
submission.

\subparagraph{Issues with the ESOP 2017 submission.}

The submitted manuscript: 

\begin{center}
  \url{https://arxiv.org/pdf/1604.08080v3.pdf}
\end{center}

\noindent
%
No specific technical criticism was raised about this version of the
paper, and the reasons for rejection (with the majority of the reviews
being positive) were given by the following summary of the PC
discussion:

\vspace{10pt}

\emph{The specifications given in the paper look rather complicated.
  The introduction dismisses linearizability, but does not give
  compelling examples showing that the more complicated specifications
  are necessary.}

\vspace{10pt}

\noindent
%
While we did not change the specifications given in Figure~4 of the
present submission from the previous version, we argued throughout the
paper that the components of the pre/postconditions from Figure~4 are
essential to reason, in a Hoare-style logic, about a large class of
interesting concurrent snapshot clients (with examples given in from
Section~4)---an aspect that neither of the works on linearizability we
are aware of considers.
%
We have added Section 8 with a discussion, comparing our specification
and proof method to those of linearizability and arguing for the
advantages of our specifications, comparing to seemingly simpler ones.
%
We have also rewritten the introduction, so it would point out
shortcomings of linearizability with respect to compositional
client-side reasoning, justifying our method that relies on auxiliary
state instead. 
%
Finally, we extended the related work with a comparison on the
concurrent work on proving linearizability using partial orders.

% \todo{Did we address this?}

% \gad{ Not yet. At least not in full. The plan is to answer them in the
%   new discussion section. However, we do not have new {\bf compelling
%     examples}. I wonder whether it is a good idea to put this summary
%   rather than listing the concerns of R1 that we do address.}

\subsection*{Source Code}

Our mechanization of Jayanti's snapshot algorithm and their clients,
together with the rest of the FCSL source files and previous case
studies is available for download from:

\begin{center}
  \url{https://software.imdea.org/~germand/relink}
\end{center}

% These files require \texttt{Coq v. 8.5.pl3} and \texttt{ssreflect
%   v.1.6}. We are working on updating them to comply with \texttt{Coq
%   v.8.6} and will upload and updated version soon.
\end{document}

% \todo{Upload a new version. The one online works on Coq 8.5.2 and
%   throws all the nasty warnings.}
