\section{Proof Outline for the Exchanger}
\label{app:exch}

\subsection{Exchanger Invariants}
%

The states in the exchanger state-space must satisfy other invariants
in addition to (\ref{tag:exchanging}). All these properties naturally
arise from our description of how the exchanger behaves on decorated
state. Thus, we simply list them here without much commentary, so that
we can refer to them further on. We abbreviate with $p \mapsto (x; y)$
the heap $p \mapsto x \hunion p\!+\!1 \mapsto y$.

\begin{enumerate}[label=(\roman*)]
\item\label{exP} $\heapj$ contains a pointer $g$ and a number of
  offers $p \mapsto (v; x)$, and $g$ points to either $\mathsf{null}$
  or to some offer in $\heapj$.

\item $\hists$, $\histo$ and $\mygather{\pending}$ contain only
  disjoint time-stamps. Similarly, $\perms$ is disjoint from $\permo$.

\item\label{matched} The offers in $\pending$ are all matched, and all
  owned by some thread:
\[
\!\!\!\!\!\!
{\small{
\exists t\ldot p \mapsto (t, v, w) \subseteq \pending \iff p \in
\perms \hunion \permo, p \mapsto (v; \Matched w) \subseteq \heapj.
}}
\]
\item There is at most one unmatched offer; it is the one linked
  from~$g$. It is owned by someone:
\[
{\small{
p \mapsto (v; \Unmatched) \subseteq \heapj \Longrightarrow p \in \perms \hunion
\permo, g \mapsto p \subseteq \heapj.
}}
\]
\item Retired offers are not owned: $p \mapsto (v; \Retired)
  \subseteq \heapj \Rightarrow p \notin \perms \hunion \permo$.

\item The outstanding offers are included in the joint heap, \ie, if
  $p \in \perms \hunion \permo$ then $p \in \mathsf{dom}\ \heapj$.

\item\label{ex:gapless} The combined history $\hists \hunion \histo \hunion
  \mygather{\pending}$ is gapless, that is, if it contains a
  time-stamp $t$, it also contains all the smaller time-stamps (sans
  0).

\end{enumerate}

{
%\setlength{\belowcaptionskip}{-10pt} 
\begin{figure}
\centering
\[
{\small{
\begin{array}{rl}
 \Num{1} & \specK{\{\heaps = \emptyset, \perms = \emptyset, \hists = \emptyset, \gist \subseteq \histo \hunion \mygather{\pending} \}}
\\ 
 \Num{2} & ~~~~ p \Asgn \esc{alloc}~(v, \Unmatched);\\
 \Num{3} & \specK{\{\heaps = p \mapsto (v; \Unmatched), \perms = \emptyset, \hists = \emptyset, \gist \subseteq \histo \hunion \mygather{\pending} \}}\\
 \Num{4} & ~~~~ b \Asgn \esc{CAS}~(g, \esc{null}, p);\\
 \Num{5} & ~~~~ \kw{if}~~b~\esc{==}~\esc{null}~~\kw{then}\\
 \Num{6} & ~~~~ ~~~~ \esc{sleep}~(50);\\
 \Num{7} & \specK{\{\heaps = \emptyset, \perms = \{p\}, \hists = \emptyset, \gist \subseteq \histo \hunion \mygather{\pending}, \esc{bounded}\ p\ v\ \gist \}}\\
 \Num{8} & ~~~~ ~~~~ x \Asgn \esc{CAS}~(p\esc{+}1, \Unmatched, \Retired);\\
 \Num{9} & \specK{\{\heaps = \emptyset, \perms = \emptyset, \gist \subseteq \histo \hunion \mygather{\pending},\hbox{}} \\
\Num{10} & \specK{\hphantom{\}}x = \Matched w \implies \exists t\ldot \hists = t \mapsto (v, w), \mathsf{last}(\gist) < t, \twin t,}\\
\Num{11} & \specK{\hphantom{\}}x = \Unmatched \implies \hists = \emptyset \}}\\
\Num{12} & ~~~~ ~~~~ \kw{if}~~x~~\kw{is}~~\Matched w~~\kw{then}~~\kw{return}~~(\esc{Some}~w)\\
\Num{13} & ~~~~ ~~~~ \kw{else}~~\kw{return}~~\esc{None}\\
\Num{14} & ~~~~ \kw{else}\\
\Num{15} & \specK{\{\heaps = p \mapsto (v; \Unmatched), \perms = \emptyset, \hists = \emptyset, \gist \subseteq \histo \hunion \mygather{\pending}\}}\\
\Num{16} & ~~~~ ~~~~ \esc{dealloc}~p;\\
\Num{17} & \specK{\{\heaps = \emptyset, \perms = \emptyset, \hists = \emptyset, \gist \subseteq \histo \hunion \mygather{\pending}\}}\\
\Num{18} & ~~~~ ~~~~ cur \Asgn \esc{read}~g;\\
\Num{19} & \specK{\{\heaps = \emptyset, \perms = \emptyset, \hists = \emptyset, \gist \subseteq \histo \hunion \mygather{\pending},}\\
\Num{20} & \specK{\hphantom{\}}cur = \esc{null} \vee cur \mapsto (w; -) \subseteq \heapj\}}\\
\Num{21} & ~~~~ ~~~~ \kw{if}~~cur~\esc{==}~\esc{null}~~\kw{then}~~\kw{return}~{\esc{None}}\\
%& \specK{\{ \mbox{obvious} \}}\\
\Num{22} & ~~~~ ~~~~ \kw{else}\\
\Num{23} & \specK{\{\heaps = \emptyset, \perms = \emptyset, \hists = \emptyset, \gist \subseteq \histo \hunion \mygather{\pending},}\\
\Num{24} & \specK{\hphantom{\}} cur \mapsto (w; -) \subseteq \heapj\}}\\
\Num{25} & ~~~~ ~~~~ ~~~~ x \Asgn \esc{CAS}(cur\esc{+}1, \Unmatched, \Matched v);\\
\Num{26} & \specK{\{\heaps = \emptyset, \perms = \emptyset, \gist \subseteq \histo \hunion \mygather{\pending}, cur \mapsto (w; y) \subseteq \heapj,} \\
\Num{27} & \specK{\hphantom{\}}x = \Unmatched \implies y = \Matched v, \exists t\ldot \hists = t \mapsto (v, w), \mathsf{last}(\gist) < t, \twin t}\\
\Num{28} & \specK{\hphantom{\}}x \neq \Unmatched \implies \hists = \emptyset, y \neq \Unmatched \}}\\
\Num{29} & ~~~~ ~~~~ ~~~~ \esc{CAS}~(g, cur, \esc{null});\\
\Num{30} & \specK{\{\mbox{same as above; the state satisfies (iv) because $y \neq \Unmatched$}\}}\\
%& \specK{\{\heaps = \emptyset, \perms = \emptyset, \gist \subseteq \histo \hunion \mygather{\pending}, 
%    \exists w hw. cur \mapsto (w, hw) \in \heapj,} \\
%& \specK{x = \Unmatched, hw = \Matched v, \exists t. \hists = t \mapsto (w, v), last \gist < t, \twin t}\\
%& \specK{x \neq \Unmatched, \hists = \emptyset, hw \neq \Unmatched \}}\\
\Num{31} & ~~~~ ~~~~ ~~~~ \kw{if}~~x~\esc{==}~\Unmatched~~\kw{then}~~w\Asgn \esc{read}~cur;\kw{return}~(\esc{Some}\ w)\\
\Num{32} & \specK{\{ \heaps = \emptyset, \perms = \emptyset, \gist \subseteq \histo \hunion \mygather{\pending}, \res=\esc{Some}\ w}, \\
\Num{33} & \specK{\hphantom{\}}\exists t. \hists = t \mapsto (w, v), \mathsf{last}(\gist) < t, \twin t\}}\\
\Num{34} & ~~~~ ~~~~ ~~~~ \kw{else}~~\kw{return}~\esc{None}\}\\
\Num{35} & \specK{\{\heaps = \emptyset, \perms = \emptyset, \hists = \emptyset, \gist \subseteq \histo \hunion \mygather{\pending}, \res=\esc{None}\}} 
\end{array}
}}
\]
%\vspace{-10pt}  
\caption{Proof outline for the exchanger.}
\label{fig:exchanger_proof}
\end{figure} 
}

\subsection{Proof outline for the exchanger}
\label{sec:exproof}

The proof outline for the spec~\eqref{tag:exchangespec} is given in
Figure~\ref{fig:exchanger_proof}, and we proceed to explain it.
%
We start with the desired precondition, and after allocation in line
2, $\heaps$ stores the offer $p$ in line 3.

If the \code{CAS} at line 4 succeeds, the program ``installs'' the
offer; that is, the state (real and auxiliary) is changed
simultaneously to the modification of $g$. In particular, $p$ is added
to $\perms$, and the offer $p$ changes ownership, to move from
$\heaps$ to $\heapj$.  Since $b$ will be bound to $\mathsf{null}$,
this leads us to the assertion in line 7. We explain in
Section~\ref{sec:background} how these kinds of changes to the
auxiliary state, which are supposed to occur simultaneously with some
atomic operation (in this case, \code{CAS}), are specified and
verified in FCSL. The assertion in line 7 further states
$\mathsf{bounded}\ p\ v\ \gist$. We do not formally define
$\mathsf{bounded}$ here (it is in the proof scripts, accompanying the
paper), but it says that $p$ has been moved to $\heapj$, \ie,
$p \mapsto (v; -) \subseteq \heapj$, and that any time-stamp $t$ at
which another thread may match $p$, and thus place the entry
$p \mapsto (t, v,-)$ into $\pending$, must satisfy
$\mathsf{last}(\gist) < t, \twin{t}$. Intuitively, this property is
valid, and stable under interference, because entries in $\pending$
can be added only by generating fresh time-stamps wrt.~the collective
history $\histo \hunion \mygather{\pending}$, and $\gist$ is a subset
of it.
%
If the \code{CAS} in line 4 fails, then nothing changes, and we simply move
to the spec in line 15.

At line 8, the \code{CAS} succeeds if $x\,{=}\,\Unmatched$, and fails if
$x\,{=}\,\Matched w$. Notice that $x$ cannot be $\Retired$; since we
own $p \in \perms$, no other thread could retire $p$.
%
If \code{CAS} fails, then the offer has been matched with~$w$. \code{CAS}
simultaneously ``collects'' the offer as follows. By invariant (iii),
and $\mathsf{bounded}\ p\ v\ \gist$, the auxiliary map $\pending$
contains an entry $p \mapsto (t, v, w)$, where $\mathsf{last}(\gist) <
t, \twin{t}$. The auxiliary state is changed to remove $p$ from
$\pending$, and simultaneously place $t \mapsto (v, w)$ into $\hists$.
%
If \code{CAS} succeeds, the offer was unmatched, and is ``retired'' by
removing $p$ from $\perms$.

Lines 12 and 13 branch on $x$, and select either the assertion 10 or
11. Either way, the desired postcondition directly follows.

%Deallocation in line 16 removes the offer $p$ from the heap $\heaps$.

After reading $cur$ in line 18, by invariant (i), we know that $cur$
either points to $\mathsf{null}$, or to some offer $p \mapsto (w; -)
\subseteq \heapj$.

%The null-check in line 21, together with line 19, directly establishes
%the postcondition.

At line 25, the \code{CAS} succeeds if $x = \Unmatched$ and fails
otherwise. If \code{CAS} succeeded, then it ``matches'' the offer in $cur$;
that is, it writes $\Matched w$ into the hole of $cur$, and changes the
auxiliary state as follows. It takes $t$ to be the smallest unused
time-stamp in the history $\hist = \hists \hunion \histo \hunion
\mygather{\pending}$. Thus $\mathsf{last}(\hist) < t$, and because
$\hist$ has even size by invariant~(\ref{tag:exchanging}), $t$ must be
odd, and hence $t < \twin t = t+1$. The $t \mapsto (v, w)$ is placed
into $\hists$, giving us assertion 27.  To preserve the invariant
(iii), \code{CAS} simultaneously puts the entry $p \mapsto (t, w, v)$ into
$\pending$, for future collection by the thread that introduced offer
$cur$. But, we do not need to reflect this in line 27.
%
If the \code{CAS} fails, the history $\hists$ remains empty, as no matching is
done. However, the hole $y$ associated with $cur$ cannot be
$\Unmatched$, as then \code{CAS} would have succeded.
%
Therefore, it is sound in line 29 to ``unlink'' $cur$ from $g$, as the
unlinking will not violate the invariant (iv), which says that an
unmatched offer must be pointed to by $g$.
%
Finally, lines 31 and 34 select the assertion 27 or 28, and either
way, directly imply the desired postcondition.



