\begin{abstract}

  We present a novel model of concurrent computations with shared
  memory and provide a simple, yet powerful, logical framework for
  uniform Hoare-style reasoning about partial correctness of coarse-
  and fine-grained concurrent programs.
%
  The key idea is to specify arbitrary resource protocols as
  communicating \emph{state transition systems} (STS) that describe
  valid states of a resource and the transitions the resource is
  allowed to make, including transfer of heap ownership.

  We demonstrate how reasoning in terms of communicating STS makes it
  easy to crystallize behavioral invariants of a resource. We also
  provide \emph{entanglement} operators to build large systems from an
  arbitrary number of STS components, by interconnecting their lines
  of communication.
%
  Furthermore, we show how the classical rules from the Concurrent
  Separation Logic (CSL), such as scoped resource allocation, can be
  generalized to fine-grained resource management. This allows us to
  give specifications as powerful as Rely-Guarantee, in a concise,
  scoped way, and yet regain the compositionality of CSL-style
  resource management.
%
  We proved the soundness of our logic with respect to the
  denotational semantics of action trees (variation on Brookes' action
  traces). We formalized the logic as a shallow embedding in Coq and
  implemented a number of examples, including a construction of
  coarse-grained CSL resources as a modular composition of various
  logical and semantic components.
\end{abstract}
